Presheaf Model for Dependent Types

### 2017-09-29 papers/talk-11/paper-11.tex
Proofread context extension model
Rewrite sections on dependent product and sum

\begin{definition}[Context extension]
  If $\vdash \Gamma~ctxt$ and $\Gamma \vdash A~type$, there is a context
  $\Gamma, x:A$ (for brevity also written as $\Gamma.A$ from now on), a context
  morphism $\mathbf{p}: \Gamma.A \to \Gamma$, and a term $\Gamma.A \vdash
  \mathbf{q} : A \mathbf{p}$. This should satisfy the following property: for
  each $\vdash \Delta~ctxt$, a context morphism $\sigma \from \Delta \to
  \Gamma$, and $\Delta \vdash a : A \sigma$, there is a context morphism
  $(\sigma, a): \Delta \to \Gamma.A$ called \emph{context extension} such that:
  \[
    \mathbf{p}(\sigma, a) = \sigma \quad \mathbf{q}(\sigma, a) = a \quad (\sigma, a) \tau =
    (\sigma \tau, a \tau) \quad (\mathbf{p}, \mathbf{q}) = 1
  \]
  Note that in the last equation $1: \Gamma.A \to \Gamma.A$ and that we leave
  the indices on $\mathbf{p}$ implicit.
\end{definition}

@@

Let us see how the category of presheaves over $\cat{C}$ (we denote it as
$Psh(\cat{C})$) forms a category of families where presheaves are contexts.  If
we set presheaves to be the contexts, then the context morphisms are morphisms
in $[\cat{C}^\op, \Set]$, which are natural transformations between contravariant
functors. These natural transformations satisfy the following naturality
condition:
\[
  \xymatrix{
    \Delta(I) \ar[r]^{\sigma_I} \ar[d]^{\Delta_f} & \Gamma(I) \ar[d]^{\Gamma_f} \\
    \Delta(J) \ar[r]^{\sigma_J} & \Gamma(J)
  }
\]
where $I, J \in \ob(\cat{C}); \Gamma, \Delta \in Psh(\cat{C}); \sigma
\from \Delta \to \Gamma$, and $f \from J \to I$. $\Gamma_f$ and $\Delta_f$
denote the morphisms that we obtain from the morphism $f$ in $\Set$ by the
functoriality of $\Gamma$ and $\Delta$. For instance, if $\rho \in \Gamma(I)$,
then we have a morphism $\Gamma_f$ to the corresponding $\rho f \in \Gamma(J)$
(we write $f$ applied on the right to connect to the initial morphism in
$\cat{C}$) since the contravariant functor $\Gamma$ preserves morphism
mappings.

@@

To define terms of a dependent type $\Gamma \vdash A~type$ we require a family
of elements $a(I, \rho) \in A(I, \rho)$ for each $I \in \ob(\cat{C})$ and $\rho
\in \Gamma(I)$. These elements should satisfy for any $f \from J \to I$ the
equation
\[
  A(f, \rho)(a(I, \rho)) = a(J, \Gamma_f \rho).
\]
The context morphism $\sigma \from \Delta \to \Gamma$ then operates on a term
$\Delta \vdash a\sigma : A\sigma$ by composition
\[
  (a\sigma_I)(I, \rho') = a(I, \sigma_I(\rho')).
\]
for $\rho' \in \Delta(I)$.

We define the presheaf $\Gamma.A$ corresponding to a context extension for
objects $I \in \ob(\cat{C})$ as
\[
  (\Gamma.A)(I) = \{(\rho, u)~|~\rho \in \Gamma(I) \land u \in A(I, \rho)\}
\]
and for morphisms $f \from J \to I$ as
\[
  (\Gamma.A)_f(\rho, u) = ((\Gamma.A)_f \rho, A(f, \rho)(u)).
\]
The projections are then defined by $\mathbf{p} \from \Gamma.A \to \Gamma$,
$\mathbf{p}(\rho, u) = \rho$, and $\Gamma.A \vdash \mathbf{q} : A\mathbf{p}$ by
$\mathbf{q}(\rho, u) = u$. Under context morphism $\sigma \from \Delta \to
\Gamma$ we define an extension of $\Gamma$ with a term $\Delta \vdash a :
A\sigma$, $(\sigma, a) \from \Delta \to \Gamma.A$ by
\[
  (\sigma, a)\rho = (\sigma \rho, a \rho)
\]
for the appropriate subscript on $\sigma$.

@@

So far we have a model for types $\Gamma \vdash A~type$ and $\Gamma.A \vdash
B~type$. Now we want to find a model for a dependent function type $\Gamma
\vdash \Pi x:A.B$. This model should connect our representations $A(I, \rho)$
and $B(\rho, u)$ for these types, where $I \in \ob(\cat{C})$, $\Gamma \in
\ob(Psh(\cat{C}))$, $\rho \in \Gamma(I)$, $A \in Psh(\int \Gamma)$, $u \in A(I,
\rho)$.

We can take that as a definition. Then an element $w$ of $(\Delta^\Gamma)(I)$
is a natural transformation $w \from \mathbf{y}I\times \Gamma \to \Delta$ given
by functions
\[
  w_J \from (\mathbf{y}I)(J) \times \Gamma(J) \to \Delta(J),
\]
and the naturality condition becomes $(w_J(f, \rho))g = w_K(fg, \rho g)$ for $f
\from J \to I$, $\rho \in \Gamma(J)$, and $g \from K \to J$.

\begin{lemma}
  CwF associated to a presheaf category supports $\Pi$-types.
  \begin{proof}
    We give a proof sketch as seen in \cite{Huber}. More detailed proof may be
    found in \cite{Hofmann}.

    Given $\Gamma \vdash A$ and $\Gamma.A \vdash B$ we have to define $\Gamma
    \vdash \Pi x: A. B~type$, that is, we have to define the set $(\Pi x: A.B)
    \rho$ for $I \in \cat{C}$ and $\rho \in \Gamma(I)$. The elements $w$ of
    $(\Pi x:A.B)\rho$ are families $w = (w_f~|~J\in \cat{C}, f \from J \to I)$
    of (dependent) functions such that
    \[
      w_f u \in B(\rho f, u) \qquad \text{for $J \in \cat{C}, f \from J \to I, and u \in A(\rho f)$},
    \]
    with the requirement that for $g \from K \to J$
    \[
      (w_f u)g = w_{fg}(ug).
    \]

    For such a family $w \in (\Pi x: A. B)\rho$, the restriction $wf \in (\Pi
    x:A.B)(\rho f)$ for $f: \from J \to I$ is defined by taking
    \[
      (w f)_g u = w_{fg}u \in B(\rho f g, u)
    \]
    where $g \from K \to J$ and $u \in A(\rho f g)$. This definition satisfies
    $w 1 = w$ and $w f g = w(f g)$.

    Now let $\Gamma.A \vdash b : B$; we have to define $\Gamma \vdash \lambda b
    : \Pi x:A.B$, that is , give $(\lambda b)\rho \in (\Pi x:A.B)\rho$. For $f
    \from J \to I$ and $u \in A(\rho f)$ we set
    \[
      ((\lambda b)\rho)_f u = b (\rho f, u) \in B(\rho f, u)
    \]
    This satisfies for $g \from K \to J$
    \[
      (((\lambda b)\rho)_f u)g = (b(\rho f, u))g = b(\rho(fg), ug) = ((\lambda b)\rho)_{fg}(ug)
    \]
    and defines a term since
    \[
      (((\lambda b)\rho)f)_g u = ((\lambda b) \rho)_{fg} u = b(\rho fg, u) = ((\lambda b)(\rho f))_g u.
    \]

    To define the application let $\Gamma \vdash u : \Pi x:A.B$ and $\Gamma
    \vdash v : A$, and we set $App(u, v)\rho = (u\rho)_1 (v\rho) \in B(\rho,
    v\rho)$, thus $App(u, v)\rho \in B(1, v)\rho$ as $(\rho, v \rho) = (1,
    v)\rho$. $\beta$-equality is readily checked
    \[
      App(\lambda b, v)\rho = ((\lambda b)\rho)_1(v\rho)=b(\rho, v\rho)
    \]
    and similarly for $\eta$-equality and other equations.
  \end{proof}
\end{lemma}

\begin{example}
  Let us consider a term of type $A \to A$, and the elements in $Ty(\Gamma)$
  and $Ter(\Gamma, A \to A)$ for some $\vdash \Gamma~ctxt$, $\Gamma \vdash
  A~type$ in the presheaf model.

  For a CwF $(\cat{C}, F), I, J \in ob(\cat{C}), f \from J \to I$ we give the
  element of $\mathit{Ty}(\Gamma)$ corresponding to the type $A \to A$:
  \[
    (A \to A)\rho = \{w | w_f u \in A(\rho f), u \in A(\rho f)\} = \{\lambda I f u.u\}
  \]
  The required equalities hold trivially for any $g \from K \to J$, $((\lambda
  I f u.u)_f u) g = (\lambda I f u.u)_{fg} (ug)$.

  $\Gamma, x : A \vdash x : A$ corresponds to $q$ in the model. Thus $\Gamma
  \vdash \lambda x.x : A -> A$ corresponds to $\lambda q$ being defined as
  follows:
  \[
    ((\lambda q) \rho)_f u = q(\rho f, u) = u
  \]
\end{example}

\subsection*{Dependent sum}
The interpretation of $\Gamma \vdash \Sigma x:A.B~type$ for $\Gamma \vdash
A~type$ and $\Gamma.A \vdash B~type$ is defined by
\[
  (\Sigma x :A. B)\rho = \{(a, b)~|~a \in A\rho \land b \in B(\rho, a)\}
\]
for $\rho \in \Gamma(I), I \in \ob(\cat{C})$.
\begin{lemma}
  CwF associated to a presheaf category supports $\Sigma$-types.
  \begin{proof}
    Again we give a proof sketch as seen in \cite{Huber}. More detailed proof
    is also found in \cite{Hofmann}.

    The restrictions are defined componentwise:
    \[
      (a, b)f = (af, bf)
    \]
    for $f \from J \to I$.

    The pairing operation $\Gamma \vdash (u, v) : \Sigma x:A.B$ of $\Gamma
    \vdash u : A$ and $\Gamma.A \vdash v : B(1, u)$ is defined by the
    componentwise pairing:
    \[
      (u, v)\rho = (u\rho, v\rho)
    \]

    Likewise for the projections $\mathbf{p}$ and $\mathbf{q}$,
    $(\mathbf{p}w)\rho = a$ and $(\mathbf{q}w)\rho = b$ for $w\rho = (a, b)$.
    This validates the necessary equations.
  \end{proof}
\end{lemma}

\section*{Credits}

Most of the material on the dependent types bases on \cite{Hofmann}, while
starting from categories with families we closely follow \cite{Huber}.

>>>>

\begin{definition}[Context extension]
  If $\vdash \Gamma~ctxt$ and $\Gamma \vdash A~type$, there is a context
  $\Gamma, x:A$ (for brevity also written as $\Gamma.A$ from now on), a context
  morphism $\mathbf{p}: \Gamma.A \to \Gamma$, and a term $\Gamma.A \vdash
  \mathbf{q} : A \mathbf{p}$. This should satisfy the following property: for
  each $\vdash \Delta~ctxt$, a context morphism $\sigma \from \Delta \to
  \Gamma$, and $\Delta \vdash a : A \sigma$, there is a context morphism
  $(\sigma, a): \Delta \to \Gamma.A$ called \emph{context extension} such that:
  \[
    \mathbf{p}(\sigma, a) = \sigma \quad \mathbf{q}(\sigma, a) = a \quad (\sigma, a) \tau =
    (\sigma \tau, a \tau) \quad (\mathbf{p}, \mathbf{q}) = 1
  \]
  Note that in the last equation $1: \Gamma.A \to \Gamma.A$.
\end{definition}

@@

Let us see how the category of presheaves over $\cat{C}$ (we denote it as
$Psh(\cat{C})$) forms a category of families where presheaves are contexts.  If
we set presheaves to be the contexts, then the context morphisms are morphisms
in $[\cat{C}^\op, \Set]$, which are natural transformations between contravariant
functors. These natural transformations satisfy the following naturality
condition:
\[
  \xymatrix{
    \Delta(I) \ar[r]^{\sigma_I} \ar[d]^{\Delta_f} & \Gamma(I) \ar[d]^{\Gamma_f} \\
    \Delta(J) \ar[r]^{\sigma_J} & \Gamma(J)
  }
\]
where $I, J \in \ob(\cat{C}); \Gamma, \Delta \in Psh(\cat{C}); \sigma
\from \Delta \to \Gamma$, and $f \from J \to I$. $\Gamma_f$ and $\Delta_f$
denote the morphisms that we obtain from the morphism $f$ in $\Set$ by the
functoriality of $\Gamma$ and $\Delta$. For instance, if $\rho \in \Gamma(I)$,
then we have a morphism $\rho \mapsto \rho f$ written as $\Gamma_f$ to the
corresponding $\rho f \in \Gamma(J)$, (we write $f$ applied on the right to
connect to the initial morphism in $\cat{C}$) since the contravariant functor
$\Gamma$ preserves morphism mappings.

@@

To define terms of a dependent type $\Gamma \vdash A~type$ we require a family
of elements $a(I, \rho) \in A(I, \rho)$ for each $I \in \ob(\cat{C})$ and $\rho
\in \Gamma(I)$. Additionally, since $A$ is a presheaf these elements should
satisfy for any $f \from J \to I$ the equation
\[
  A(f, \rho)(a(I, \rho)) = a(J, \Gamma_f \rho).
\]
The context morphism $\sigma \from \Delta \to \Gamma$ then operates on a term
$\Delta \vdash a\sigma : A\sigma$ by composition
\[
  (a\sigma_I)(I, \rho') = a(I, \sigma_I(\rho')).
\]
for $\rho' \in \Delta(I)$.

We define the presheaf $\Gamma.A$ corresponding to a context extension for
objects $I \in \ob(\cat{C})$ as
\[
  (\Gamma.A)(I) = \{(\rho, u)~|~\rho \in \Gamma(I) \land u \in A(I, \rho)\}
\]
and for morphisms $f \from J \to I$ as
\[
  (\Gamma.A)_f(\rho, u) = ((\Gamma.A)_f \rho, A(f, \rho)(u)).
\]
The projections are then defined by $\mathbf{p} \from \Gamma.A \to \Gamma$,
$\mathbf{p}(\rho, u) = \rho$, and $\Gamma.A \vdash \mathbf{q} : A\mathbf{p}$ by
$\mathbf{q}(\rho, u) = u$. Under context morphism $\sigma \from \Delta \to
\Gamma$ we define an extension of $\Gamma$ with a term $\Delta \vdash a :
A\sigma$, $(\sigma, a) \from \Delta \to \Gamma.A$ by
\[
  (\sigma, a)(I, \rho') = (\sigma_I \rho', a(I, \rho'))
\]

@@

So far we have a model for types $\Gamma \vdash A~type$ and $\Gamma.A \vdash
B~type$. Now we want to find a model for a dependent function type $\Gamma
\vdash \Pi x:A.B$ that should connect a term $u \in A(I, \rho)$ with a term $v
\in B(I, (\rho, a))$. In other words, the element $\pi$ of $(\Pi x:A.B)(I, \rho)$
should satisfy
\[
  \pi(u) \in B(I, (\rho, u))
\]
for some $I \in \ob(\cat{C})$, $\Gamma \in \ob(Psh(\cat{C}))$, $\rho \in
\Gamma(I)$.

However, this condition alone is not enough: since $\Pi x: A.B$ is a presheaf
we also need to preserve the equalities for the morphisms in $\int \Gamma$
(which are morphisms in $\cat{C}$). To generalize for the morphism $f \from J
\to I$ (where $I, J \in \ob(\cat{C}))$) we extend $\pi \in (\Pi x: A.B)(I,
\rho)$ to a family of functions $\pi = (\pi_f~|~f \from J \to I)$ and require
the following equation to hold:
\[
  \pi_f(u) \in B(J, (\Gamma_f \rho, A(f, \rho)(u)))
\]
where $\rho \in \Gamma(I)$ and $u \in A(J, \Gamma_f \rho)$. Notice that since
$(f, \rho)$ is a morphism in $\int \Gamma$ we have $A(f, \rho)(u) \in A(I,
\Gamma_f \rho)$. Another point is that if we take $1 \from I \to I$ instead of
$f$, then we get our above condition immediately.

We also need to respect compositions:
\[
  \pi_{fg}(A(g, \Gamma_f \rho)(u)) = B(g, (\Gamma_f \rho, A(f, \rho)(u)))(\pi_f(u))
\]
for some $g \from K \to J$.

And define the necessary equality for operations of the morphisms in $\cat{C}$
on the right
\[
  ((\Pi u: A.B)(h, \Gamma_f \rho)(\pi))_f(u) = \pi_{hf}(u)
\]
where $h \from I' \to I$, $\pi \in (\Pi x:A.B)(I, \rho)$, $f \from J \to I'$,
$u \in A(J, \Gamma_f(\Gamma_h \rho))$, and $(\Pi x: A.B)(h, \Gamma_f \rho)(\pi)
\in (\Pi x: A.B')(I, \Gamma_h \rho)$.

Let us now define terms $\lambda x: A. b$ of a dependent function type. If we
have some $b \in B(I, (\rho, x))$, then we define $(\lambda x: a.b)(I, \rho)
\in (\Pi x: A.B)(I, \rho)$ by:
\[
  ((\lambda u: A.b)(I, \rho))_f(u) = b(J, (\Gamma_f \rho, u)) \in B(J, (\Gamma_f \rho, u))
\]
where $f \from J \to I$ and $u \in A(J, \Gamma_f \rho)$. Note that this
definition satisfies the required equalities above.

What remains is to define the application of the lambda terms. First let
$\Gamma \vdash u : \Pi x:A.B$ and $\Gamma \vdash v : A$. The application
morphism $App$ is then thus defined for some $I \in \ob(\cat{C})$:
\[
  App(u, v)(I, \rho) = (u(I, \rho))_1(v(I, \rho)) \in B(I, (\rho, v(I, \rho)))
\]

Since the necessary elements of $\Pi$-types are in place, we can state the soundness lemma:
\begin{lemma}
  CwF associated to a presheaf category supports $\Pi$-types.
  \begin{proof}
    Follows after the verification of the $\Pi$-rules on the above equations. A
    detailed proof is found $\cite{Hofmann}$
  \end{proof}
\end{lemma}

\begin{example}
  Let us consider a representation of a term of type $A \to A$ for some $\vdash
  \Gamma~ctxt$, $\Gamma \vdash A~type$ in the presheaf model. For a CwF $(\cat{C}, F), I, J \in ob(\cat{C}), f \from J \to I$ we give the
  element of $\mathit{Ty}(\Gamma)$ corresponding to the type $A \to A$:
  \[
    (A \to A)\rho = \{\pi~|~\pi_f(u) \in A(\rho f), u \in A(\rho f)\} = \{\lambda I f u.u\}
  \]
  Note that the required equalities hold trivially for any $g \from K \to J$, $((\lambda
  I f u.u)_f u) g = (\lambda I f u.u)_{fg} (ug)$. $\Gamma, x : A \vdash x : A$ corresponds to $q$ in the model. Thus $\Gamma
  \vdash \lambda x.x : A -> A$ corresponds to $\lambda q$ being defined as
  follows:
  \[
    ((\lambda q) \rho)_f u = q(\rho f, u) = u
  \]
\end{example}

\subsection*{Dependent sum}
Suppose we have some types $\Gamma \vdash A~type$ and $\Gamma.A \vdash B~type$.
Then we can define the dependent pair type $\Gamma \vdash \Sigma x:A.B~type$
for some $I \in \ob(\cat{C})$ by
\[
  (\Sigma a :A. B)(I, \rho) = \{(a, b)~|~a \in A(I, \rho) \land b \in B(I, (\rho, a))\}
\]
where $\rho \in \Gamma(I)$.

Note that as is the case with the dependent product we need to take the
morphisms in $\cat{C}$ into account if we want $\Sigma x:A.B$ to be a presheaf.
This leads us to the following equation for some $f \from J \to I$:
\[
  \Gamma_f(a, b) = (\Gamma_f a, \Gamma_f b)
\]

Also, for a pair to be a term, both of its elements should terms in the
following manner:
\[
  (u, v)(I, \rho) = (u(I, \rho), v(I, \rho))
\]
where $u \in A(I, \rho)$ and $v \in B(I, (\rho, u))$.

Lastly, we need projection morphisms to be defined:
\[
  (\mathbf{p}(w))(I, \rho) = a
\]
\[
  (\mathbf{q}(w)(I, \rho) = b
\]
where $w(I, \rho) = (a, b)$.

\begin{lemma}
  CwF associated to a presheaf category supports $\Sigma$-types.
  \begin{proof}
    Follows after verification of $\Sigma$-rules. A detailed proof is found in
    \cite{Hofmann}.
  \end{proof}
\end{lemma}
